---
filters:
  - naquiz
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this workshop:"
---

```{r, include=FALSE}
library(magrittr)
library(knitr)
library(kableExtra)
options(readr.show_col_types = FALSE)
```

# RNA-seq workshop

Which genes are differentially expressed between three cell types in the mouse mammary gland? In this workshop we answer that question for basal, luminal progenitor (LP) and mature luminal (ML) cells. The cells were sorted from the mammary glands of female virgin mice, with three replicates per cell type (nine samples), and sequenced on Illumina RNA-seq platforms. The data comes from the `RNAseq123` workflow.

By the end you will have a table of differentially expressed (DE) genes for each pairwise comparison, an MA plot and a heatmap of the top DE genes, all saved to an `output` folder.

To get there, we follow the [limma-voom workflow](https://f1000research.com/articles/5-1408):

- filtering out lowly expressed genes
- normalisation
- creating a multidimensional scaling (MDS) plot
- creating a design matrix
- fitting gene-wise linear models, with empirical Bayes moderation to better estimate gene-wise variability
- testing for differential expression

The focus is on running the workflow and building the data structures the packages need, rather than on the underlying statistics, which we cover only as far as needed to interpret the results. For more detail, see the `limma` User's Guide by typing `help("limma")` and following the links.

## Data files

You can download and unzip the files using the following commands:

```{r}
download.file("https://github.com/Shians/BaseR_Intro/raw/master/data/workshop_data.zip", "workshop_data.zip")
unzip("workshop_data.zip", exdir = "data")
unzip("data/Ses3_counts.zip", exdir = "data/counts")
```

## R packages

The packages required for this workshop can be installed with the following commands:

```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RNAseq123")
```

## Loading in read count data and annotation

We will use the `edgeR` package to load and store the data.

To use the `edgeR` package we first need to construct a `DGEList` object. This object contains three key pieces of data:

- `counts`: the main data of this object, a matrix of count values with samples along the columns and features/genes along the rows.
- `samples`: a data frame containing annotation for the samples. The rows in this table describe the corresponding column of the counts data.
- `genes`: a data frame containing annotation for the genes in the counts matrix. The rows in this table describe the corresponding row in the counts matrix.

::: {.callout-note title="Data objects"}

Data objects in general are a way to store related pieces of information together. Often they provide features to maintain the relationships between the pieces of information. For example, in a `DGEList` object, the `counts` are stored as a matrix while `samples` and `genes` are stored as data frames. When you subset the `DGEList` object, the rows and columns of the `counts` matrix are subset along with the corresponding rows of the `samples` and `genes` data frames. This prevents errors that can occur if the user tries to manually subset all three pieces of information separately.

:::

```{r, warning=FALSE, message=FALSE}
# load required packages
library(tidyverse)
library(edgeR)

group <- parse_factor(
  c("LP", "ML", "Basal", "Basal", "ML", "LP", "Basal", "ML", "LP"),
  levels = c("Basal", "LP", "ML")
)

samplenames <- c(
  "10_6_5_11", "9_6_5_11", "purep53", "JMS8-2", "JMS8-3",
  "JMS8-4", "JMS8-5", "JMS9-P7c", "JMS9-P8c"
)
```

::: {.callout-note collapse="true" title="A brief detour to factors"}
When we declared the `group` variable, we used a factor. Factors are a special data type used to encode categorical variables. They have several useful properties compared with regular character vectors:

- They allow you to specify the order of the levels.
- They are stored as integers but displayed as labels.
- They encode all the valid levels of the factor, even if they are not present in the data.

Specifying the order of the levels is useful because it allows you to rearrange labels when used in plots. In this data we would have the "Basal" group first, followed by "LP" and then "ML". Using `parse_factor()` also allows you to check that the values are all valid levels. For example, if one of the samples were labelled "Bassal" instead of "Basal", it would throw an error. You can read the R for Data Science [chapter on factors](https://r4ds.had.co.nz/factors.html) for more information, and see the [forcats package](https://forcats.tidyverse.org) for many useful functions for working with factors.
:::

## Creating the DGEList object

We will create a `DGEList` object following the `RNAseq123` workflow. We use the `readDGE()` function to read in count data from files, provide sample information to `readDGE()` at the same time, and add the gene annotation information afterwards.

```{r, warning=FALSE, message=FALSE}
# vector of file names
files <- dir(path = "data/counts", pattern = "GSM*")

# create DGEList object
dge <- readDGE(
  files,
  path = "data/counts",
  columns = c(1, 3),
  group = group,
  labels = samplenames
)

# add gene annotation information
gene_anno <- read_tsv("data/Ses3_geneAnnot.tsv")
dge$genes <- gene_anno
```

## Alternative construction of the DGEList object

A more common way to create a `DGEList` object is to read in the count data as a matrix and then create the `DGEList` object using the `DGEList()` function. We can pull apart our existing `DGEList` object and recreate it using this method. In practice, you will likely be downloading count data as a matrix from a file, and then creating the `DGEList` object using the `DGEList()` function, so this is a useful exercise to understand how to construct the data object from its components.

::: {.callout-note}
What you can do with data objects is determined by the package that the object comes from. You will need to read the package documentation to find out how to access data from the object. Here `edgeR` informs us that the `DGEList` has data that can be accessed as if it were a list.
:::

```{r}
counts <- dge$counts    # Usually read in from some counts.txt file
samples <- dge$samples  # Usually read in or constructed from sample metadata
genes <- dge$genes      # Usually read in from some gene_anno.txt or constructed from rownames of the counts matrix

# create DGEList object
dge <- DGEList(counts = counts, samples = samples, genes = genes)
```

Now we have a data object that can be used for downstream analysis.

## Filtering

The first step of our analysis is to filter out lowly expressed genes. There are two main problems with low-abundance genes:

- Technical variation is more problematic for low-abundance genes. This variation is thought to be due to two factors: insufficient mixing and low sampling fraction [@mcintyre2011].
  - Insufficient mixing of solutions during library preparation can result in uneven distribution of reads.
  - RNA-seq can be thought of as sampling. Measurement errors will occur simply due to the random nature of the sampling process. This problem affects low-abundance RNA species more because the relative error for small count values is larger than it would be for more highly abundant RNA species.
- Genes that are very lowly expressed do not produce sufficient information to be useful for biological interpretation. For example, it is very hard to believe the biological significance of genes that have counts ranging from 0 to 3 across samples even if they come up as statistically significant.

Filtering out lowly expressed genes means fewer statistical tests, so less multiple testing correction and more power to detect DE genes [@bourgon2010]. 'Power' is your ability to detect true positives. DE analysis runs one test per gene, and with thousands of tests some will be significant purely by chance. 'Multiple testing correction' adjusts the p-values to control these false positives, at the cost of reduced power. The fewer tests we perform, the smaller that cost.

The `edgeR` package provides the `filterByExpr()` function to automate gene filtering. By default, it aims to keep genes with a count of 10 or more, in at least as many samples as the smallest experimental group. In our experiment, there are three phenotype groups, each with three samples. Therefore we retain only genes that have 10 or more counts in three or more samples. The actual filtering is done on counts per million (cpm), to prevent bias against samples with small library sizes. This complex procedure is the reason why the package provides a function to perform the filtering for you.

The output of this function is a vector of logicals, indicating which genes (rows) should be kept and which should be filtered out.

```{r}
keep <- filterByExpr(dge)
table(keep)
proportions(table(keep))

dge <- dge[keep, , keep.lib.sizes = FALSE] # drop the filtered genes and recalculate library sizes
dim(dge$counts)
```

We can see that we now have 16624 genes. We started with 27179 genes, meaning that ~40% of genes have been filtered out.

## Library-size normalisation

After filtering, our next step is to normalise the data. Normalisation refers to the process of adjusting the data to reduce or eliminate systematic bias. This allows the data to be meaningfully compared across samples or experimental groups.

There are two main factors that need to be normalised for in RNA-seq:

- Sequencing depth/library size: technically, sequencing a sample to half the depth will give, on average, half the number of reads mapping to each gene [@robinson2010].
- RNA composition: if a large number of genes are unique to, or highly expressed in, only one experimental condition, the sequencing capacity available for the remaining genes in that sample is decreased. For example, suppose only five genes are studied in two experimental groups. If one gene is particularly high in group A, then with limited sequencing depth that gene will reduce the counts of the remaining four genes. As a result, the remaining four genes appear under-expressed in group A compared to group B, even though the true amount of gene product is equal for these four genes [@robinson2010].

Sequencing depth is accounted for by calculating the cpm. This metric is calculated by:

1.  taking the library size (sum of all counts for a sample),
2.  dividing this by 1,000,000 to get the 'per million' scaling factor,
3.  then dividing all read counts for each gene in that sample by the 'per million' scaling factor

RNA composition can be accounted for by using more sophisticated normalisation methodologies. We will use 'trimmed mean of M-values' (TMM), which estimates relative RNA levels from RNA-seq data [@robinson2010]. Under the assumption that most genes are not differentially expressed, TMM calculates a library size scaling factor for each library (sample) from the genes that remain after trimming those with the most extreme log-fold-changes and expression levels.

::: {.callout-tip collapse="true" title="Going deeper: how TMM works"}
TMM calculates the scaling factor for each library using the following steps:

1.  calculate the gene expression log-fold-changes and absolute expression values for pairwise samples (selecting one sample from the experiment as a reference)
2.  remove the genes with the highest and lowest log-fold-changes and absolute expression values
3.  take a weighted mean of the remaining genes (where the weight is the inverse of the approximate asymptotic variances). This gives the normalisation factor for each library (sample)
:::

Subsequent steps in this analysis will use log-cpm values, calculated using the normalisation factors to scale each library size.

We can calculate the normalisation factors, specifying that we want to use the `"TMM"` method:

```{r}
dge <- calcNormFactors(dge, method = "TMM")
```

This function calculates the normalisation factors for each library (sample) and puts this information in the `samples` data frame. Note that it takes `dge` (our `DGEList` object) as input and returns a `DGEList` object as well.

Let's take a look at our normalisation factors:

```{r}
dge$samples
```

These normalisation factors are close to 1 for all samples, suggesting minimal difference in RNA composition between samples.

## Visualising the effect of normalisation

To see the effect of TMM normalisation, we plot each sample's distribution of log-counts as a boxplot, before and after normalisation. `ggplot2` needs the data in long format, with one row per observation, so we first reshape the count matrix.

```{r}
# log-transform the raw counts (adding 0.5 avoids taking the log of zero),
# then convert the matrix to a tibble, keeping row names as a "gene" column
as_tibble(log(dge$counts + 0.5), rownames = "gene") %>%
  # reshape from wide to long: one row per gene-sample pair
  pivot_longer(
    where(is.numeric),
    names_to = "sample", values_to = "expression"
  ) %>%
  ggplot(aes(x = sample, y = expression)) +
  geom_boxplot()
```

We can compare this to the cpm values. The `cpm()` function automatically applies normalisation factors if they are present.

```{r}
as_tibble(cpm(dge, log = TRUE), rownames = "gene") %>%
  pivot_longer(
    where(is.numeric),
    names_to = "sample", values_to = "expression"
  ) %>%
  ggplot(aes(x = sample, y = expression)) +
  geom_boxplot()
```

We see that by performing normalisation on our data, the gene expression values of each sample now have similar medians and quantiles. This indicates that the relative expression values of each sample can be more meaningfully compared.

## MDS plots

Before we perform statistical tests, it's useful to perform some exploratory visual analysis to get an overall idea of how our data is behaving.

MDS is a way to visualise distances between sets of data points (samples in our case). It is a dimensionality reduction technique, similar to principal components analysis (PCA). We treat gene expression in samples as if they were coordinates in a high-dimensional coordinate system. We can then find "distances" between samples as we do between points in space. The goal of the algorithm is to find a representation in lower-dimensional space such that the distance between any two points in high-dimensional space is preserved in lower dimensions.

The `plotMDS()` function from `limma` creates an MDS plot from a `DGEList` object.

```{r}
plotMDS(dge)
```

Each point on the plot represents one sample and is 'labelled' using the sample name. The distances between samples in the resulting plot can be interpreted as the typical log2-fold-change between the samples, for the most differentially expressed genes.

We can change the labelling to use the name of the group the sample belongs to instead:

```{r}
plotMDS(dge, labels = group)
```

This shows us that the phenotype groups tend to cluster together, meaning that the gene expression profiles are similar for samples within a phenotype group. The 'Basal' type samples are quite close together while the 'LP' (luminal progenitor) and 'ML' (mature luminal) type samples are further apart, signifying that their expression profiles are more variable.

```{r}
# load required packages
library(tidyverse)
library(edgeR)
library(limma)

group <- parse_factor(
  c("LP", "ML", "Basal", "Basal", "ML", "LP", "Basal", "ML", "LP"),
  levels = c("Basal", "LP", "ML")
)

samplenames <- c(
  "10_6_5_11", "9_6_5_11", "purep53", "JMS8-2", "JMS8-3",
  "JMS8-4", "JMS8-5", "JMS9-P7c", "JMS9-P8c"
)

# create DGEList object
dge <- readDGE(
  dir(path = "data/counts", pattern = "GSM*"),
  path = "data/counts",
  columns = c(1, 3),
  group = group,
  labels = samplenames
)

# add gene annotation information
dge$genes <- read_tsv("data/Ses3_geneAnnot.tsv")

# filter lowly expressed genes
keep <- filterByExpr(dge)
dge <- dge[keep, , keep.lib.sizes = FALSE]

# normalise the data for library size and RNA composition
dge <- calcNormFactors(dge, method = "TMM")
```

## Linear modelling

To identify differentially expressed genes from our normalised gene expression data, we will use the `limma` package to fit linear models to the genes. Linear models are a broad class of statistical models that fit the value of a response (or dependent) variable as a linear function of one or more explanatory (or independent) variables (also called covariates).

The general form of a linear model looks like this:

$Y = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2}... + \epsilon$

This equation is saying that a response variable of interest $Y$ is equal to a constant ($\beta_{0}$) plus the sum of the covariates ($X_{i}$) each multiplied by a constant coefficient ($\beta_{i}$) and some error term ($\epsilon$). The error term is the difference between the observed value and the predicted value of the model and is assumed to have a normal distribution with mean 0 and some variance.

Our experiment is quite simple, since there is only a single covariate, the cell type. The true benefit of using linear models is in their ability to accommodate more complex designs including multiple covariates.

To fit the linear models in the limma-voom framework we need two objects in addition to our data:

-   A design matrix, representing the covariates.
-   A contrast matrix, representing the specific comparison we wish to make.

### Design matrix

The first step to fitting a linear model is to specify a design matrix. The design matrix specifies the values of the covariates for each sample, and is represented as a matrix for mathematical convenience.

To generate a design matrix, we use the function `model.matrix()`, with the expression `~0 + group`. This returns a matrix representing the design where there is no intercept term and group is the only covariate. If we omit the `0` then there would be an intercept in the model, and if we included more covariates then more columns would be generated.

```{r}
design <- model.matrix(~0 + group, data = dge$samples)
design
```

There are nine rows, one for each sample. Along the columns are the names of the groups. The values in the cells denote membership of the particular sample for a particular group. As our groups in this case are mutually exclusive, each row contains only a single 1 to denote membership in a single group.

### Contrasts

Contrast matrices are companions to design matrices. They are used to specify the comparison of interest. In our case, we have three experimental groups: `Basal`, `LP` and `ML`. So if we are to perform differential expression analysis, we are most likely interested in the differences between only two of the groups at a time. Contrast matrices let us specify which comparison we'd like to make, and are also represented as a matrix just like the design.

A contrast matrix can be made using the `makeContrasts()` function. Within this function, we specify the name of each specific contrast and the formula for that contrast. For example, the `BasalvsLP` contrast compares the difference between the `Basal` and `LP` groups. Note that the names of the phenotype groups must be written exactly as they are in the column names of our design matrix (see above).

In addition to the individual contrasts, the function must know about the design of the model. This is passed through the `levels` argument, which either accepts a matrix with the column names corresponding to levels of your experimental groups, or the levels themselves as a character vector. Here we first simplify the column names of our design matrix to make it easier to read. Then we create the contrast matrix using the `makeContrasts()` function.

```{r}
colnames(design) <- c("Basal", "LP", "ML")

contr_matrix <- makeContrasts(
  BasalvsLP = "Basal - LP",
  BasalvsML = "Basal - ML",
  LPvsML = "LP - ML",
  levels = design # alternatively 'levels = colnames(design)'
)

contr_matrix
```

**The order of the contrast sets the sign: for `Basal - LP`, a positive log-fold-change means higher expression in Basal.** Had we written `LP - Basal`, genes up-regulated in Basal would instead have a negative log-fold-change.

::: {.callout-tip collapse="true" title="Going deeper: why the columns sum to zero"}
The numbers in each column of a contrast matrix sum to 0. Each contrast compares groups against each other, so the weights on one side balance those on the other.
:::

### Variance modelling with voom

We are now ready to fit our linear models. `limma` fits linear models to the data with the assumption that the underlying data is normally distributed. Count data is generally not normally distributed, but log transforming count data gives it a roughly normal distribution sufficient for linear models to work well. To do this, `limma` transforms the raw count data to log-cpm using library sizes and the normalisation factors we calculated previously.

In addition to the normalisation steps, the limma-voom pipeline uses the `voom()` function to generate weights for the individual observations based on a modelled mean-variance relationship. The mean-variance relationship is used to estimate the variance of each individual observation of sample-gene expression, and by applying these weights to the linear modelling step, it helps the data better meet the assumptions of the linear model and results in more accurate statistical inferences.

The `voom()` function takes as arguments our `DGEList` object and our design matrix. It also optionally outputs a plot of the mean-variance relationship of our data, called the 'voom plot'.

```{r}
v <- voom(dge, design, plot = TRUE)
```

The voom plot shows the mean-variance relationship of the data. Each point is a gene: the x-axis is its average log2 expression, and the y-axis is the square root of its residual standard deviation from the linear model ("sqrt(standard deviation)"). The red line is the trend fitted across all genes. The idea is that the variance of a gene can be estimated as a function of its mean expression, and that this estimate is more accurate than the direct calculation of the variance.

A good voom plot shows a smooth trend that decreases as expression increases: lowly expressed genes are more variable than highly expressed ones. If the trend instead rises sharply at the low-expression end, many very lowly expressed genes remain, which suggests filtering was insufficient.

The output of `voom()` (our variable `v`) is an `EList` object which contains the following elements:

-   `genes`: a data frame of gene annotation data.
-   `targets`: a data frame of sample data.
-   `E`: a numeric matrix of normalised log-cpm values.
-   `weights`: a numeric matrix of precision weights.
-   `design`: the design matrix.

### Fitting the linear model

We are now ready to fit our linear model with `lmFit()`, which estimates the coefficients defined in our design matrix `design`. The resulting object, `vfit`, is an `MArrayLM` object. It contains information about our genes (the same data frame as `genes` from our `EList` object `v` above), the design matrix and a number of statistical outputs. Of most interest to us are the coefficients, stored in an element called `coefficients`. The first rows of this matrix are shown below. Each gene is a row and is labelled using its Entrez ID. Each column gives coefficients for each of our phenotype groups. These coefficients are weighted averages of the log-cpm of each gene in each group.

```{r}
vfit <- lmFit(v)
head(vfit$coefficients)
```

We can then use `contrasts.fit()` to calculate coefficients for each contrast (or 'comparison') we specified in our `contr_matrix`. The output is also an `MArrayLM` object. When we inspect the `coefficients` element now, we can see that each column is a contrast that we specified in our contrast matrix.

```{r}
vfit <- contrasts.fit(vfit, contrasts = contr_matrix)
head(vfit$coefficients)
```

A difference significantly greater than 0 means the gene is up-regulated in the first group of the contrast; significantly less than 0 means it is down-regulated. If the difference is not significantly different from 0, we have failed to establish a difference in expression between the two groups.

## Statistical testing

To actually test if the values are significantly different from 0, we use the `eBayes()` function to compute a moderated t-statistic for each gene, given a fitted linear model. 'Moderated' refers to empirical Bayes moderation, which borrows information across genes to obtain more accurate measures of variability for each gene. This also increases our power to detect differentially expressed genes.

::: {.callout-tip collapse="true" title="Going deeper: other statistics from eBayes()"}
As well as moderated t-statistics, `eBayes()` computes moderated F-statistics, which test all contrasts at once, and the log-odds of differential expression for each gene. The log-odds appears as the `B` column in the output of `topTable()`.
:::

```{r}
efit <- eBayes(vfit)
```

We can now look at the number of differentially expressed genes using the `decideTests()` function. The output of this function is a matrix where each column is a contrast (comparison of interest) and each row is a gene. The numbers 1, -1 and 0 mean up-regulated, down-regulated or not significantly differentially expressed, respectively. The statistical significance is determined by the adjusted p-value being less than 0.05.

By default, `decideTests()` adjusts p-values for multiple testing using the Benjamini and Hochberg [@benjamini1995] method, but several others are also available.

```{r}
dt <- decideTests(efit)
head(dt)
```

To obtain the total number of differentially expressed genes for each comparison, we can add the function `summary()`:

```{r}
summary(dt)
```

Both comparisons involving the basal cells have a large number of DE genes (more than 9,000 each), while LP vs ML has the fewest (about half as many). This matches the MDS plot, where the basal samples separate from the LP and ML samples along the first dimension, and the LP and ML samples sit closer to each other.

The function `topTable()` can be used to obtain more information on the differentially expressed genes for each contrast. `topTable()` takes as arguments the `MArrayLM` object output by `eBayes()` (`efit`), the contrast name of interest and the number of top differentially expressed genes to output. Note that the contrast name must be given in quotes and must be exactly as written in the contrast matrix `contr_matrix`.

It outputs a data frame with the following information:

-   **Gene details**: gene information, from the `genes` element of the `MArrayLM` object (`efit`).
-   `logFC`: the log2-fold-change of the contrast.
-   `AveExpr`: the average log2 expression of that gene.
-   `t`: the moderated t-statistic.
-   `P.Value`: the p-value.
-   `adj.P.Val`: the adjusted p-value.
-   `B`: log-odds that the gene is differentially expressed.

```{r}
top <- topTable(efit, coef = "BasalvsLP", n = Inf)
head(top)
```

By default the genes are sorted by the `B` statistic, so the genes with the strongest evidence of DE come first. Because the contrast is `Basal - LP`, a positive `logFC` means the gene is higher in basal cells and a negative `logFC` means it is higher in LP cells. The six genes shown here all have negative `logFC`, so they are more highly expressed in LP cells.

## MA plot

The MA plot is a plot of log-fold-change (M-values) against log-expression averages (A-values). This is a common plot in RNA-seq analysis to visualise the result of differential expression tests. It can be created using the `plotMA()` function from the `limma` package. Creating this plot requires three pieces of information:

-   `object = efit`: the fitted object containing the log-fold-change and log-expression averages.
-   `coef = 1`: the column number of the contrast to plot since there are three different contrasts fitted within the object.
-   `status = dt[, "BasalvsLP"]`: a vector of numerics denoting whether a gene is up-regulated or down-regulated.

```{r}
plotMA(efit, coef = 1, status = dt[, "BasalvsLP"])
```

The coloured points are the DE genes: up-regulated genes (higher in basal cells) sit above zero and down-regulated genes (higher in LP cells) sit below it. Notice that at the low-expression end (left of the plot), genes rarely reach significance even with a large log-fold-change. Counts for these genes are small and noisy, so a large log-fold-change there is weak evidence of a real difference.

## Heatmap

Finally we can create a heatmap of the differentially expressed genes. The heatmap is a useful way to visualise the expression of the differentially expressed genes across the samples. We can use the `pheatmap` package to create the heatmap.

If you haven't already, install the package:

```{r, eval = FALSE}
install.packages("pheatmap")
```

We take the top 50 DE genes from each contrast, combine them, and plot their log-cpm values, labelled with gene symbols.

```{r}
# combine the top 50 genes from each contrast, removing duplicates
top_genes <- bind_rows(
  topTable(efit, coef = "BasalvsLP", n = 50),
  topTable(efit, coef = "BasalvsML", n = 50),
  topTable(efit, coef = "LPvsML", n = 50)
) %>%
  select(ENTREZID, SYMBOL, TXCHROM) %>%
  distinct()

# extract log-cpm values for these genes. ENTREZID is numeric, so convert it
# to character to select rows by ID rather than by position
top_gene_expr <- cpm(dge, log = TRUE)[as.character(top_genes$ENTREZID), ]
# label rows with gene symbols for readability
rownames(top_gene_expr) <- top_genes$SYMBOL
```

We draw the heatmap with `pheatmap()`, scaling each gene to a z-score so that patterns across samples are comparable between genes. Try removing each option to see how it affects the heatmap, and explore further options in the `pheatmap()` documentation.

```{r}
library(pheatmap)

# sample annotation: colour bar showing each sample's group
sample_anno <- dge$samples %>%
  select(group)

heatmap <- pheatmap(
  top_gene_expr,
  color = colorRampPalette(c("blue", "white", "red"))(100), # custom colour scale
  scale = "row", # z-score each gene across samples
  annotation_col = sample_anno, # add sample group annotation
  border_color = NA,
  cluster_rows = TRUE, # cluster genes
  cluster_cols = TRUE, # cluster samples
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 6 # small font so gene labels fit
)
```

The samples cluster by cell type, so the three replicates of each group sit next to each other. The genes form blocks that are high in one group and low in the others, reflecting the genes that distinguish each cell type. Because each row is scaled to a z-score, the colour shows whether a gene is high or low relative to its own average across samples, not its absolute expression level: a red cell in one row and a red cell in another do not mean the two genes are expressed at similar levels.

## Saving all the results

Now that we have produced a series of plots and the top genes table, we can save these files to a directory. These figures may form the basis for publication figures, and the results table can be partially presented as a result, or the full table can be included as a supplementary file.

```{r, warning=FALSE}
dir.create("output")

write_csv(top, "output/top_genes_BasalvsLP.csv")
ggsave("output/heatmap.png", heatmap$gtable, height = 10, width = 7)
```

With that we have come to the conclusion of the RNA-seq analysis and this course. You have now learned how to use various `tidyverse` packages to operate on data and create publication-quality figures. You have also learned how to use the `limma` and `edgeR` packages to perform differential expression analysis on RNA-seq data. You should now be able to apply these skills to perform analysis on your own data.

# Extra content

## MDS plot with ggplot2

For more customisability and better consistency with the style of other plots, it'd be nice to be able to draw the MDS plot using `ggplot2`. In order to do this we would need the MDS coordinates calculated by `plotMDS()`. Luckily the documentation of `plotMDS()` shows that it returns multiple computed values.

```{r}
mds_result <- plotMDS(dge, plot = FALSE)
mds_result
```

We see that there are `x` and `y` values in the list returned by the `plotMDS()` function. We can put these into a table and see if they match up to the positions plotted by the function.

```{r}
mds_tibble <- tibble(
  x = mds_result$x,
  y = mds_result$y
)

ggplot(mds_tibble, aes(x = x, y = y)) +
  geom_point()
```

The positions of the points and the range of the scales seem to match, so we want to add in more metadata to help us plot. We can try to recreate the MDS plot using just `ggplot2`.

```{r}
mds_tibble <- tibble(
  x = mds_result$x,
  y = mds_result$y,
  sample = colnames(dge),
  group = dge$samples$group
)

dim1_var_explained <- round(mds_result$var.explained[1] * 100)
dim2_var_explained <- round(mds_result$var.explained[2] * 100)

ggplot(mds_tibble, aes(x = x, y = y)) +
  geom_text(aes(label = group)) +
  labs(
    x = paste0("Leading logFC dim 1 ", "(", dim1_var_explained, "%)"),
    y = paste0("Leading logFC dim 2 ", "(", dim2_var_explained, "%)")
  )
```

Now that we have our data in a nice tidy format we can use with `ggplot2`, it's easy to create variations of the plot. For example we can draw points instead of group labels, and use colour to identify the groups.

```{r}
ggplot(mds_tibble, aes(x = x, y = y, col = group)) +
  geom_point() +
  labs(
    x = paste0("Leading logFC dim 1 ", "(", dim1_var_explained, "%)"),
    y = paste0("Leading logFC dim 2 ", "(", dim2_var_explained, "%)")
  )
```

Alternatively we can also use the labels to identify the individual samples while colouring them by their group.

```{r}
ggplot(mds_tibble, aes(x = x, y = y, col = group)) +
  geom_text(aes(label = sample)) +
  labs(
    x = paste0("Leading logFC dim 1 ", "(", dim1_var_explained, "%)"),
    y = paste0("Leading logFC dim 2 ", "(", dim2_var_explained, "%)")
  )
```

We see that some labels are very hard to read because they overlap, so we can use the `ggrepel` package to fix this. The `ggrepel` package creates labels that repel each other in order to avoid overlap. It's a good idea to use it in conjunction with `geom_point()` in order to keep track of exact coordinates of the data points.

If you haven't already, install the package:

```{r, eval = FALSE}
install.packages("ggrepel")
```

```{r}
library(ggrepel)
ggplot(mds_tibble, aes(x = x, y = y, col = group)) +
  geom_point() +
  geom_text_repel(aes(label = sample)) +
  labs(
    x = paste0("Leading logFC dim 1 ", "(", dim1_var_explained, "%)"),
    y = paste0("Leading logFC dim 2 ", "(", dim2_var_explained, "%)")
  )
```

## MA plot with ggplot2

We can try to recreate this plot using `ggplot2` to give us more flexibility in plotting. The information we'll need is the log-fold-change, the average log-expression and the differential expression status of each gene. We can extract this information from the `efit` and `dt` objects we created above, as well as gene names for annotation purposes.

```{r}
ma_plot_data <- tibble(
  logFC = efit$coefficients[, "BasalvsLP"],
  AvgExpr = efit$Amean,
  status = factor(as.numeric(dt[, "BasalvsLP"])),
  gene = efit$genes$SYMBOL,
  p_value = efit$p.value[, "BasalvsLP"]
)
```

With our tibble prepared, we can use `ggplot2` to create the plot as we have done before:

```{r}
ma_plot_data %>%
  ggplot(aes(x = AvgExpr, y = logFC, col = status)) +
  geom_point()
```

While the coordinates of the points are the same, there is still quite a bit of work to do to make this plot look like the one produced by `plotMA()`. First, we fix the colours. Second, we draw the non-significant points in their own layer underneath, so they don't hide the DE genes. Finally, we label the 15 most significant genes with `ggrepel`.

```{r}
ma_plot_data_fct <- ma_plot_data %>%
  mutate(
    status = fct_recode(
      status,
      "Not DE" = "0",
      "Down" = "-1",
      "Up" = "1"
    )
  ) %>%
  mutate(status = fct_relevel(status, "Not DE", "Down", "Up"))

top_de_genes <- ma_plot_data_fct %>%
  arrange(p_value) %>%
  slice_head(n = 15)

scale_cols <- c(
  "Not DE" = "black",
  "Down" = "blue",
  "Up" = "red"
)

ma_plot <- ma_plot_data_fct %>%
  ggplot(aes(x = AvgExpr, y = logFC, col = status)) +
  geom_point(data = ma_plot_data_fct %>% filter(status == "Not DE"), size = 0.1) +
  geom_point(data = ma_plot_data_fct %>% filter(status != "Not DE"), size = 1) +
  geom_label_repel(
    data = top_de_genes,
    aes(label = gene, col = NULL),
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = scale_cols)

ma_plot

ggsave("output/ma_plot_BasalvsLP.png", ma_plot)
```

## Volcano plot

Since we have the tibble of the data prepared, we can also choose to create an alternative plot called a volcano plot, which is also very popular in RNA-seq analysis. The volcano plot is a scatter plot of the negative log10 p-value against the log2-fold-change. This plot is useful for visualising the significance of the differentially expressed genes.

```{r}
volcano_plot <- ma_plot_data_fct %>%
  ggplot(aes(x = logFC, y = -log10(p_value), col = status)) +
  geom_point(data = ma_plot_data_fct %>% filter(status == "Not DE"), size = 0.1) +
  geom_point(data = ma_plot_data_fct %>% filter(status != "Not DE"), size = 1) +
  geom_label_repel(
    data = top_de_genes,
    aes(label = gene, col = NULL),
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = scale_cols)

volcano_plot

ggsave("output/volcano_plot_BasalvsLP.png", volcano_plot)
```


## References
